---
title: "Analyzing an epidemic with SEIR and SIR dynamics"
output: pdf_document
---

In this vignette, we will demonstrate how to implement the Particle Marginal Metropolis-Hastings (PMMH) algorithm (which has been implemented in the pomp package) for analyzing prevalence counts from an outbreak of influenza in a British boarding school. This dataset was pulled from the pomp package, and was described and analyzed in Fintzi, Cui, Wakefield, and Minin (2016).

#Influenza in a British boarding school (SIR model)

We begin by read the dataset.

```{r,echo=T,eval=T}
library(pomp)
library(ggplot2)
library(gtools)
set.seed(52787)
st.date <- as.Date("1978/1/22")
en.date <- as.Date("1978/2/4")
date.seq <- seq(st.date,en.date, by = "day")

bbs <- data.frame(time = rep(date.seq,1),
                  count = c(rep("Observed count", 14)),
                  I = c(3,8,28,76,222,293,257,237,192,126,70,28,12,5))
```

We use the SIR model to model the British Boarding data here with a binomial sampling distribution. The hosts are divided into three classes, according to their status. The susceptible class (S) contains those that have not yet been infected and are thereby still susceptible to it; the infected class (I) comprises those who are currently infected and, by assumption, infectious; the removed class (R) includes those who are recovered or quarantined as a result of the infection. Individuals in R are assumed to be immune against reinfection. It is natural to formulate this model as a continuous-time Markov process. To start with, we need to specify the measurement model.

```{r,echo=T,eval=T}
rmeas<-"
cases=rbinom(I,exp(rho)/(1+exp(rho)));//represent the data
"
dmeas<-"
lik=dbinom(cases,I,exp(rho)/(1+exp(rho)),give_log); //return the loglikelihood
"
```

Here, we are using the $\textbf{cases}$ to refer to the data (number of reported cases) and $\textbf{I}$ to refer to the true incidence over the reporting interval. The binomial simulator rbinom and density function dbinom are provided by R. Notice that, in these snippets, we never declare the variables; pomp will ensure tht the state variable ($\textbf{I}$), observables ($\textbf{cases}$), parameters ($\textbf{rho}$) and likelihood ($\textbf{lik}$) are defined in the contexts within which these snippets are executed.

And for the transition of the SIR model, we use a tau-leaping algorithm, one version of which is implemented in the pomp package via the $\bf{euler.sim}$ plug-in. The algorithm holds the transition rates constant over a small interval of time and simulates the numbers of transitions that occur over the interval. The functions $\textbf{reulermultinom}$ draw random deviates from such distributions. Then we need to first specify a function that advances the states from $t$ to $t+\triangle t$, and we give the trantition of the SIR model code below: 

```{r,echo=T,eval=T}
sir.step<-"
double rate[2];
double dN[2];
rate[0]=exp(bet)*I;    //Infection rate
rate[1]=exp(mu);       //recovery rate
reulermultinom(1,S,&rate[0],dt,&dN[0]);   //generate the number of newly infected people
reulermultinom(1,I,&rate[1],dt,&dN[1]);   //generate the number of newly recovered people
if(!R_FINITE(S)) Rprintf(\"%lg %lg %lg %lg %lg %lg %lg %lg %lg\\n\",dN[0],rate[0],dN[1],rate[1],bet,mu,S,I,R);
S+=-dN[0];          //update the number of Susceptible
I+=dN[0]-dN[1];     //update the number of Infection
R+=dN[1];           //update the number of Recovery
" 
```

The pomp will ensure that the undeclared state variables and parameters are defined in the context within which the snippet is executed. And the $\textbf{rate}$ and $\textbf{dN}$ arrays hold the rates and numbers of transition events, respectively. Then we could construct the pomp objects below:

```{r,echo=T,eval=T}
sir <- pomp(
  data = data.frame(cases = bbs[,3], time = seq(0, 13, by = 1)),  #"cases" is the dataset, "time" is the observation time
  times = "time",
  t0 = 0,  #initial time point
  dmeasure = Csnippet(dmeas),  
  rmeasure = Csnippet(rmeas),  #return the likelihood
  rprocess = euler.sim(step.fun = Csnippet(sir.step), delta.t = 1/12), #delta.t is here
  statenames = c("S", "I", "R"),  #state space variable name
  paramnames = c("bet", "mu", "rho", "theta1", "theta2"), #parameters name
  initializer = function(params, t0, ...) {
    ps <-exp(params["theta1"]) / (1 + exp(params["theta1"]) + exp(params["theta2"]))  #given the initial proportion of susceptible
    pi <- 1 / (1 + exp(params["theta1"]) + exp(params["theta2"]))                     #given the initial proportion of infection
    pr <-exp(params["theta2"]) / (1 + exp(params["theta1"]) + exp(params["theta2"]))  #given the initial proportion of recovery
    return(setNames(as.numeric(rmultinom(
      1, 763, prob = c(ps, pi, pr)
    )), c("S", "I", "R")))
  },
  params = c(    #for data generation(We didn't generate data here, but pomp need this part.)
    bet = -5,  
    mu = -1,
    rho = 3,
    theta1 = 5,
    theta2 = 1
  )
)
```

To carry out Bayesian inference we need to specify a prior distribution on unknown parame- ters. The pomp constructor function provides the rprior and dprior arguments, which can be filled with functions that simulate from and evaluate the prior density, respectively. Methods based on random walk Metropolis-Hastings require evaluation of the prior density (dprior), so we specify dprior for the SIR model as follows.

```{r,echo=T,eval=T}
sir.dprior <- function(params, ..., log) {
  f <-
    (
      dgamma(
        exp(params[1]),
        shape = 0.001,
        rate = 1,
        log = TRUE
      ) + params[1] +          #log prior for log(infection rate) "beta"
        dgamma(
          exp(params[2]),
          shape = 1,
          rate = 2,
          log = TRUE
        ) + params[2] +        #log prior for log(recovery rate) "mu"
        dbeta(exp(params[3]) / (1 + exp(params[3])), 2, 1, log = TRUE) +
        params[3] - log((1 + exp(params[3])) ^ 2) + #log prior for logit(sampling probablity) "rho"
        log(ddirichlet(c(
          exp(params[4]) / (1 + exp(params[4]) + exp(params[5])),
          1 / (1 + exp(params[4]) + exp(params[5])),
          exp(params[5]) / (1 + exp(params[4]) + exp(params[5]))
        ), c(900, 3, 9))) + params[4] + params[5] - 3 * log(1 + exp(params[4]) +
                                                              exp(params[5])) #log prior for logit(initial value
    )
  if (log)
    f
  else
    exp(f)
} 
```

And starting value of the PMMH is specified below:

```{r,echo=T,eval=T}
param.initial <- c(  #initial value of the parameters of the PMCMC
  bet = -6,
  mu = -1,
  rho = 2.5,
  theta1 = 6,
  theta2 = 1
)
```

And the following runs 1 PMMH chain for 2000 tunning steps (using 5000 particles) to get a suitable multivariate normal random walk proposal diagonal variance-covariance matrix value.

```{r,echo=T,eval=T,warning=F}
pmcmc1 <- pmcmc(
          pomp(sir, dprior = sir.dprior),
          #given the prior function
          start = param.initial,
          #given the initial value of the parameters
          Nmcmc = 10,
          #number of mcmc steps
          Np = 5000,
          max.fail = Inf,
          proposal = mvn.rw.adaptive(0.01 * c(
                    bet = 0.1,
                    #sampling variance for beta
                    mu = 0.1,
                    #sampling variance for mu
                    rho = 0.3,
                    #sampling variance for rho
                    theta1 = 0.2,
                    #sampling variance for theta1
                    theta2 = 0.2 #sampling variance for theta2
          ),
          shape.start = 100, 
          scale.start = 100,
          max.scaling = 1.2)
)
```

And the following runs 1 PMMH chain for 100000 steps (using 5000 particles) with suitable multivariate normal random walk proposal diagonal variance-covariance matrix value we get above to get the posterior distribution of the interested parameters.

```{r,echo=T,eval=T,warning=F}
start_time <- Sys.time();  #calculation of time
pmcmc1 <- pmcmc(
          pmcmc1,
          #given the prior function
          start = param.initial,
          #given the initial value of the parameters
          Nmcmc = 10,
          max.fail = Inf,
          proposal = mvn.rw(covmat(pmcmc1))
)

end_time <- Sys.time(); #calculation of time
run_time <- difftime(end_time, start_time, units = "hours") #calculation of time
pomp_results <- list(time = run_time, results = pmcmc1) 
```

#Influenza in a British boarding school (SEIR model)

Here, we use the SEIR model to model the British Boarding data here with a binomial sampling distribution. The hosts are divided into four classes, according to their status. The susceptible class (S) contains those that have not yet been infected and are thereby still susceptible to it; the exposure class (E) contains those that have not yet been infected but already exposed to it; the infected class (I) comprises those who are currently infected and, by assumption, infectious; the removed class (R) includes those who are recovered or quarantined as a result of the infection. Individuals in R are assumed to be immune against reinfection. It is natural to formulate this model as a continuous-time Markov process. To start with, we need to specify the measurement model.

```{r,echo=T,eval=T}
rmeas<-"
cases=rbinom(I,exp(rho)/(1+exp(rho)));//represent the data
"
dmeas<-"
lik=dbinom(cases,I,exp(rho)/(1+exp(rho)),give_log); //return the loglikelihood
"
```

Here, we are using the $\textbf{cases}$ to refer to the data (number of reported cases) and $\textbf{I}$ to refer to the true incidence over the reporting interval. The binomial simulator rbinom and density function dbinom are provided by R. Notice that, in these snippets, we never declare the variables; pomp will ensure tht the state variable ($\textbf{I}$), observables ($\textbf{cases}$), parameters ($\textbf{rho}$) and likelihood ($\textbf{lik}$) are defined in the contexts within which these snippets are executed.

And for the transition of the SEIR model, we use a tau-leaping algorithm, one version of which is implemented in the pomp package via the $\bf{euler.sim}$ plug-in. The algorithm holds the transition rates constant over a small interval of time and simulates the numbers of transitions that occur over the interval. The functions $\textbf{reulermultinom}$ draw random deviates from such distributions. Then we need to first specify a function that advances the states from $t$ to $t+\triangle t$, and we give the trantition of the SEIR model code below: 

```{r,echo=T,eval=T}
seir.step<-"
double rate[3];
double dN[3];
rate[0]=exp(beta)*I;   //Infection
rate[1]=exp(gamma);    //Rate of S to E
rate[2]=exp(mu);       //Rate of E to I
reulermultinom(1,S,&rate[0],dt,&dN[0]);   //generate the number of newly from S to E
reulermultinom(1,E,&rate[1],dt,&dN[1]);   //generate the number of newly from E to I
reulermultinom(1,I,&rate[2],dt,&dN[2]);   //generate the number of newly from I to R
if(!R_FINITE(S)) Rprintf(\"%lg %lg %lg %lg %lg %lg %lg %lg %lg\\n\",dN[0],rate[0],dN[1],rate[1],beta,mu,S,I,R);
S+=-dN[0];       //update the number of Susceptible
E+=dN[0]-dN[1];  //update the number of E
I+=dN[1]-dN[2];  //update the number of I
R+=dN[2];        //update the number of R
"
```

The pomp will ensure that the undeclared state variables and parameters are defined in the context within which the snippet is executed. And the $\textbf{rate}$ and $\textbf{dN}$ arrays hold the rates and numbers of transition events, respectively. Then we could construct the pomp objects below:

```{r,echo=T,eval=T}
seir <- pomp(
  data = data.frame(cases = bbs[,3], time = seq(0, 13, by = 1)),  #"cases" is the dataset, "time" is the observation time
  times = "time",
  t0 = 0,  #initial time point
  dmeasure = Csnippet(dmeas),  
  rmeasure = Csnippet(rmeas),  #return the likelihood
  rprocess = euler.sim(step.fun = Csnippet(seir.step), delta.t = 1/12), #delta.t is here
  statenames = c("S", "E", "I", "R"), #state space variable name
  paramnames = c("beta", "gamma", "mu", "rho", "theta1", "theta2", "theta3"), #parameters name
  initializer = function(params, t0, ...) {
            #Initial proportion of S, E, I, R
            ps <- exp(params["theta1"]) / (1 + exp(params["theta1"]) + exp(params["theta2"]) + exp(params["theta3"]))
            pe <- exp(params["theta2"]) / (1 + exp(params["theta1"]) + exp(params["theta2"]) + exp(params["theta3"]))
            pi <- 1 / (1 + exp(params["theta1"]) + exp(params["theta2"]) + exp(params["theta3"]))
            pr <- exp(params["theta3"]) / (1 + exp(params["theta1"]) + exp(params["theta2"]) + exp(params["theta3"]))  
            return(setNames(as.numeric(rmultinom(1, 763, prob = c(ps, pe, pi, pr))), c("S", "E", "I", "R")))
  },
  params = c(    #for data generation(We didn't generate data here, but pomp need this part.)
    beta = -5,
    gamma = -1,
    mu = -1,
    rho = 3,
    theta1 = 5,
    theta2 = 1,
    theta3 = 1
  )
)
```

To carry out Bayesian inference we need to specify a prior distribution on unknown parame- ters. The pomp constructor function provides the rprior and dprior arguments, which can be filled with functions that simulate from and evaluate the prior density, respectively. Methods based on random walk Metropolis-Hastings require evaluation of the prior density (dprior), so we specify dprior for the SEIR model as follows.

```{r,echo=T,eval=T}
trans<-function(a,b,c){
          a_t<-exp(a)
          b_t<-exp(b)
          c_t<-exp(c)
          return(a_t*b_t*c_t/((1+a_t+b_t+c_t)^4))
}

seir.dprior <- function(params, ..., log) {
          f <- (dgamma(exp(params[1]), 0.001, 1, log = TRUE) + params[1] +  #log prior for beta
                          dgamma(exp(params[2]), 0.001, 1, log = TRUE) + params[2] +  #log prior for gamma
                          dgamma(exp(params[3]), 1, 2, log = TRUE) + params[3] +   #log prior for mu
                          dbeta(exp(params[4]) / (1 + exp(params[4])), 2, 1, log = TRUE) +
                          params[4] - log((1 + exp(params[4])) ^ 2) +      #log prior for rho
                          log(ddirichlet(c(exp(params[5]) / (1 + exp(params[5]) + exp(params[6]) + exp(params[7])),
                                           exp(params[6]) / (1 + exp(params[5]) + exp(params[6]) + exp(params[7])),
                                           1 / (1 + exp(params[5]) + exp(params[6]) + exp(params[7])),
                                           exp(params[7]) / (1 + exp(params[5]) + exp(params[6]) + exp(params[7]))),
                                         c(900, 6, 3, 9)) * trans(params[5], params[6], params[7]))  #log prior for initial SEIR value
          )
          if (log) {
                    f
          } else {
                    exp(f)
          }
}
```

And starting value of the PMMH is specified below:

```{r,echo=T,eval=T}
param.initial <- c(  #initial value of the parameters of the PMCMC
  beta = -6,
  gamma = 1.5,
  mu = -1,
  rho = 2.5,
  theta1 = 6,
  theta2 = 1,
  theta3 = 1
)
```

And the following runs 1 PMMH chain for 2000 tunning steps (using 5000 particles) to get a suitable multivariate normal random walk proposal diagonal variance-covariance matrix value.

```{r,echo=T,eval=T,warning=F}
pmcmc1 <- pmcmc(
          pomp(seir, dprior = seir.dprior),
          #given the prior function
          start = param.initial,
          #given the initial value of the parameters
          Nmcmc = 10,
          #number of mcmc steps
          Np = 5000,
          max.fail = Inf,
          proposal = mvn.rw.adaptive(0.01 * c(
                    beta = 0.1, #sampling variance for beta
                    gamma = 0.1, # sampling variance for gamma
                    mu = 0.1, #sampling variance for mu
                    rho = 0.3, #sampling variance for rho
                    theta1 = 0.2, #sampling variance for theta1
                    theta2 = 0.2, #sampling variance for theta2
                    theta3 = 0.3
          ),
          shape.start = 100, 
          scale.start = 100,
          max.scaling = 1.2)
)
```

And the following runs 1 PMMH chain for 100000 steps (using 5000 particles) with suitable multivariate normal random walk proposal diagonal variance-covariance matrix value we get above to get the posterior distribution of the interested parameters.

```{r,echo=T,eval=T,warning=F}
start_time <- Sys.time();  #calculation of time
pmcmc1 <- pmcmc(
          pmcmc1,
          #given the prior function
          start = param.initial,
          #given the initial value of the parameters
          Nmcmc = 10,
          max.fail = Inf,
          proposal = mvn.rw(covmat(pmcmc1))
)

end_time <- Sys.time(); #calculation of time
run_time <- difftime(end_time, start_time, units = "hours") #calculation of time

pomp_results <- list(time = run_time, results = pmcmc1) 
```

