---
title: "Effect of prior specification on posterior inference"
author: "Jon Fintzi, Xiang Cui, Jon Wakefield, Vladimir Minin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Additional simulations included Fintzi, Wakefield, and Minin (2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r load_packages, echo = F, include=F}
library(BDAepimodel)
library(coda)
library(Rcpp)
```

This vignette contains code to reproduce the simulation examining the prior strength in the fourth simulation of Fintzi et al. (2016). Additional details on the use of the BDAepimodel package and how to extract the results from fitted objects are provided in the "BDAepimodel" vignette. 

## Simualting the epidemic
We simulated an epidemic with SIR dynamics in a population of 750 individuals, 3% of whom were initially infected, and 7% of whom had prior immunity. The per-contact infectivity rate was $\beta = 0.00035$ and the mean infectious period duration was $1/\mu = 7$ days, which combined for a basic reproductive number of $R_0 = 1.837$. We fit four SIR models to binomially distributed weekly prevalence data, sampled with detection probability $ \rho = 0.2$, under the following four prior regimes: Regime 1 --- informative priors for all model parameters; Regime 2 --- vague priors for the rate parameters and an informative prior for the sampling probability; Regime 3 --- informative priors for the rate parameters and a flat prior for the sampling probability; Regime 4 --- vague priors for the rate parameters and a flat prior for the sampling probability. 

The data are simulated as follows:

```{r SIR_sim, warning=F, cache = F}
library(Rcpp)
library(BDAepimodel)
set.seed(52787)

# declare the functions for simulating from and evaluating the log-density of the measurement process
r_meas_process <- function(state, meas_vars, params){
          # in our example, rho will be the name of the binomial sampling probability parameter.
          # this function returns a matrix of observed counts
          rbinom(n = nrow(state), 
                 size = state[,meas_vars],
                 prob = params["rho"])
}

d_meas_process <- function(state, meas_vars, params, log = TRUE) {
          # note that the names of the measurement variables are endowed with suffixes "_observed" and "_augmented". This is required.
          # we will declare the names of the measurement variables shortly.
          dbinom(x = state[, "I_observed"], 
                 size = state[, "I_augmented"], 
                 prob = params["rho"], log = log)
}

# initialize the stochastic epidemic model object
epimodel <- init_epimodel(obstimes = seq(0, 105, by = 7),                             # vector of observation times
                          popsize = 750,                                              # population size
                          states = c("S", "I", "R"),                                  # compartment names
                          params = c(beta = 0.00035,                                  # infectivity parameter
                                     mu = 1/7,                                        # recovery rate
                                     rho = 0.2,                                       # binomial sampling probability
                                     S0 = 0.9, I0 = 0.03, R0 = 0.07),                 # initial state probabilities
                          rates = c("beta * I", "mu"),                                # unlumped transition rates
                          flow = matrix(c(-1, 1, 0, 0, -1, 1), ncol = 3, byrow = T),  # flow matrix
                          meas_vars = "I",                                            # name of measurement variable
                          r_meas_process = r_meas_process,                            # measurement process functions
                          d_meas_process = d_meas_process)

# simulate the epidemic and the dataset.  
epimodel <- simulate_epimodel(epimodel = epimodel, lump = TRUE, trim = TRUE)
plot(x = epimodel$pop_mat[,"time"], y = epimodel$pop_mat[,"I"], xlim = c(0,85), "l", xlab = "Time", ylab = "Prevalence")
points(x = epimodel$dat[,"time"], y = epimodel$dat[,"I"])
```

The next step is to define a transition kernel for the model parameters. The parameters are updated from their univariate full conditional distributions via Gibbs sampling (prior distributions presented in the code below). The prior regimes were set using an external batch function. The following code implements the transition kernel and a helper function for computing the sufficient statistics:

```{r SIR_kernel, warning = F, cache=F}
# define the hyperprior parameters for the rates and sampling probability
beta_prior <- matrix(c(3, 10000, 0.3, 1000), nrow = 2); colnames(beta_prior) <- c("informative", "diffuse")
mu_prior   <- matrix(c(3, 20, 0.1, 0.8), nrow = 2); colnames(mu_prior) <- c("informative", "diffuse")
rho_prior  <- matrix(c(21, 75, 1, 1), nrow = 2); colnames(rho_prior) <- c("informative", "diffuse") 

# set the prior for this vignette - these were set with an external batch script
rates_prior <- "informative"
samp_prior  <- "informative"

# helper function for computing the sufficient statistics for the SIR model rate parameters
Rcpp::cppFunction("Rcpp::NumericVector getSuffStats(const Rcpp::NumericMatrix& pop_mat, const int ind_final_config) {
                  
          // initialize sufficient statistics
          int num_inf = 0;       // number of infection events
          int num_rec = 0;       // number of recovery events
          double beta_suff = 0;  // integrated hazard for the infectivity
          double mu_suff = 0;    // integrated hazard for the recovery

          // initialize times
          double cur_time = 0;              // current time
          double next_time = pop_mat(0,0);  // time of the first event
          double dt = 0;                    // time increment
          
          // compute the sufficient statistics - loop through the pop_mat matrix until
          // reaching the row for the final observation time
          for(int j = 0; j < ind_final_config - 1; ++j) {
          
                    cur_time = next_time;         
                    next_time = pop_mat(j+1, 0); // grab the time of the next event
                    dt = next_time - cur_time;   // compute the time increment
                    
                    beta_suff += pop_mat(j, 3) * pop_mat(j, 4) * dt; // add S*I*(t_{j+1} - t_j) to beta_suff
                    mu_suff += pop_mat(j, 4) * dt;                   // add I*(t_{j+1} - t_j) to mu_suff
                    
                    // increment the count for the next event
                    if(pop_mat(j + 1, 2) == 1) {  
                              num_inf += 1;
                    } else if(pop_mat(j + 1, 2) == 2) {
                              num_rec += 1;
                    }
          }
                  
          // return the vector of sufficient statistics for the rate parameters
          return Rcpp::NumericVector::create(num_inf, beta_suff, num_rec, mu_suff);
}")

# MCMC transition kernel for the SIR model rate parameters and the binomial
# sampling probability. The prior distributions for the parameters are contained
# in this function.

gibbs_kernel <- function(epimodel) {
          
          # get sufficient statistics using the previously compiled getSuffStats function (above)
          suff_stats          <- getSuffStats(epimodel$pop_mat, epimodel$ind_final_config)
          
          # update parameters from their univariate full conditional distributions
          # Priors: beta ~ gamma(0.3, 1000)
          #         mu   ~ gamma(1, 8)
          #         rho  ~ beta(21, 75)
          proposal          <- epimodel$params # params is the vector of ALL model parameters
          proposal["beta"]  <- rgamma(1, beta_prior[1,rates_prior] + suff_stats[1], beta_prior[2,rates_prior] + suff_stats[2])
          proposal["mu"]    <- rgamma(1, mu_prior[1,rates_prior] + suff_stats[3], mu_prior[2,rates_prior] + suff_stats[4])
          proposal["rho"]   <- rbeta(1, 
                                     shape1 = rho_prior[1,samp_prior] + sum(epimodel$obs_mat[, "I_observed"]),
                                     shape2 = rho_prior[2,samp_prior] + sum(epimodel$obs_mat[, "I_augmented"] - 
                                                                                    epimodel$obs_mat[, "I_observed"]))
          
          # update array of rate matrices
          epimodel          <- build_new_irms(epimodel, proposal)
          
          # update the eigen decompositions (This function is built in)
          buildEigenArray_SIR(real_eigenvals = epimodel$real_eigen_values,
                              imag_eigenvals = epimodel$imag_eigen_values,
                              eigenvecs      = epimodel$eigen_vectors, 
                              inversevecs    = epimodel$inv_eigen_vectors, 
                              irm_array      = epimodel$irm, 
                              n_real_eigs    = epimodel$n_real_eigs, 
                              initial_calc   = FALSE)
          
          # get log-likelihood of the observations under the new parameters
          obs_likelihood_new  <- calc_obs_likelihood(epimodel, params = proposal, log = TRUE) #### NOTE - log = TRUE
          
          # get the new population level CTMC log-likelihood
          pop_likelihood_new  <- epimodel$likelihoods$pop_likelihood_cur +
                    suff_stats[1] * (log(proposal["beta"]) - log(epimodel$params["beta"])) +
                    suff_stats[3] * (log(proposal["mu"]) - log(epimodel$params["mu"])) -
                    suff_stats[2] * (proposal["beta"] - epimodel$params["beta"]) - 
                    suff_stats[4] * (proposal["mu"] - epimodel$params["mu"])
          
          # update parameters, likelihood objects, and eigen decompositions
          epimodel  <-
                    update_params(
                              epimodel,
                              params = proposal,
                              pop_likelihood = pop_likelihood_new,
                              obs_likelihood = obs_likelihood_new
                    )
          
          return(epimodel)
}
```

We now initialize an epimodel object with the dataset, set the RNG seed, and run each MCMC chain as follows. Note that the value for chain (chain = 1,2,3) was set by an external batch script. 

```{r SIR_inference, warning=F, cache=F, messages = F}
chain <- 1 # set by an external script

set.seed(52787 + chain)

# initial values for initial state parameters
init_dist <- MCMCpack::rdirichlet(1, c(9,0.5,0.1))
epimodel <- init_epimodel(popsize = 750,                                                       # population size
                          states = c("S", "I", "R"),                                           # compartment names
                          params = c(beta = abs(rnorm(1, 0.00035, 5e-5)),                      # per-contact infectivity rate
                                     mu = abs(rnorm(1, 1/7, 0.02)),                            # recovery rate
                                     rho = rbeta(1, 21, 75),                                   # binomial sampling probability
                                     S0 = init_dist[1], I0 = init_dist[2], R0 = init_dist[3]), # initial state probabilities
                          rates = c("beta * I", "mu"),                                         # unlumped transition rates
                          flow = matrix(c(-1, 1, 0, 0, -1, 1), ncol = 3, byrow = T),           # flow matrix
                          dat = dat,                                                           # dataset
                          time_var = "time",                                                   # name of time variable in the dataset
                          meas_vars = "I",                                                     # name of measurement var in the dataset
                          initdist_prior = c(9,0.2,0.5), ### Parameters for the dirichlet prior distribution for the initial state probs
                          r_meas_process = r_meas_process,
                          d_meas_process = d_meas_process)

epimodel <- init_settings(epimodel,
                          niter = 10, # set to 100000 for the paper
                          save_params_every = 1, 
                          save_configs_every = 250, 
                          kernel = list(gibbs_kernel),
                          configs_to_redraw = 5, # set to 75 for the paper
                          analytic_eigen = "SIR")

epimodel <- fit_epimodel(epimodel, monitor = FALSE)

```

After running all three chains for each prior regime, we discarded the burn-in and combined the parameter samples and latent posterior samples from each chain. Posterior median estimates and 95% credible intervals of model parameters were computed, along with the pointwise posterior distribution of the latent process. These are presented in the paper and the supplement. 
